From Coast to Coast: Uncovering the Provincial and Municipal impact of Wealth on the Spatial Distribution of Private and Public School Institutions in Canada

### 2023-04-20 outputs/paper/paper.qmd
Edits

---
author: "Sofia Sellitto"
date: "April 21, 2023"
date-format: long
format: 
  pdf: 
    documentclass: article 
    geometry: margin = 1in 
    abstract: "This paper follows an examination of the spatial distribution of educational facilities in Canada and whether a relationship exists between facility location and Province/municipality wealth. The research aims to identify any significant patterns and trends in the distribution of educational facilities, particularly in relation to province and municipality wealth. I have found that _________findings________. The findings of this study aim to contribute to a deeper understanding of educational access and inequality in Canada to inform future research and policies aimed at addressing these issues."
toc: true
fig_caption: yes
nocite: '@*'
bibliography: references.bib
---
```{r setup}
#| echo: false
#| warning: false
#| include: false

library(readr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(janitor)
library(knitr)
library(ggmap)
library(gridtext)
library(gridExtra)
library(sp)
library(rgdal)
library(sf)
library(marginaleffects)

@@

data_cad <- read_csv(here::here("inputs/data/data.csv"))

```{r}
#| echo: false
#| warning: false
#| label: table_1

@@

@table_1 illustrates the numerical difference between each of the provinces of interest and their associated number of school types. It is apparent that Ontario maintains the largest number of schools total across all the Provinces of interest. This distribution appears accurate as Ontario has the largest population among all the provinces in Canada. Ontario also has the highest numerical values of Private schools, with 1206. Given that Ontario has both the largest population among all Canadian provinces and is the richest province with extreme wealth disparity, the observed distribution appears to be accurate.  Alberta maintains the second highest variation of public to private schools with Nova Scotia being the third.

```{r}
#| echo: false
#| warning: false
#| label: fig_1

@@

data_cad |>
  count(Prov_Terr, Facility_Type) |>
  ggplot(aes(x = Prov_Terr, y = n, fill = Facility_Type))+
  geom_col(position = "dodge")+
  theme_minimal()+
  labs(
  title = "The Number of School Types for Each Canadian Province or Territory",
  x = "Province",
  y = "Number of Schools",
  fill = "School Type")

```
@fig_1 is a visual distribution of Table 1 to better aid in the comprehension of the data. Here the ratio of private to public schools is far more digestible. It also becomes apparent that Ontario appears to be the only Province with Catholic school. This is due to the fact that Ontario is the only province that has a separate publicly funded Catholic school system. In other provinces and territories, these types of educational institutions would be classified as private schools.

```{r}
#| echo: false
#| warning: false
#| label: map_1

@@

 geom_point(data = data_cad %>% filter(Facility_Type != "Private"), aes(x = Longitude, y = Latitude, color = Facility_Type), size = 1, alpha = 0.2) +

  geom_point(data = data_cad %>% filter(Facility_Type == "Private"), aes(x = Longitude, y = Latitude, color = Facility_Type), size = 1, alpha = 0.5) +
  scale_color_manual(values = c("green", "orange", "purple", "red", "pink")) +
  labs(
    title = "Spatial Distribution of Types of Schools Across Canada", 
    color = "Facility Type") +
  theme_minimal()

```
@map_1 showcases the spatial distribution of school types across the map of Canada. In turning your attention toward the red dots, which denote public schools,  It is visually evident from the map that public schools are concentrated in the highest volume in Ontario. Upon observing the distribution of school types across Ontario, and given its status as the richest province with the highest income inequality, I have chosen to further examine this distribution within Ontario.

```{r}
#| echo: false
#| warning: false
#| label: map_2

@@

bbox <- c(left = -98, bottom = 41.80, right = -74.471, top = 52.92)
ontario_map <- get_stamenmap(bbox,
                             zoom = 7,
                             maptype = "terrain")
#Plotting the map 
ggmap(ontario_map,
      extent = "nomral",
      maprange = FALSE) +
annotate("text", x = -85, y = 49.5, label = "Ontario", size = 3, fontface = "bold", color = "black") +

  geom_point(data = ontario_schools %>% filter(Facility_Type != "Private"), aes(x = Longitude, y = Latitude, color = Facility_Type), size = 1, alpha = 0.2) +

  geom_point(data = ontario_schools %>% filter(Facility_Type == "Private"), aes(x = Longitude, y = Latitude, color = Facility_Type), size = 1, alpha = 0.5) +
scale_color_manual(values = c("green", "red", "pink")) +
  labs(
    title = "Spatial Distribution of Types of Schools Across Ontario", 
    color = "Facility Type") +
  theme_minimal()

```
@map_2 again is a spatial distribution of school types across the map of Ontario. Again using the same colours to denote the school types, it is apparent here that the highest volume of private schools appears to cluster in Northern Ontario among the Greater Toronto Area (GTA), Ottawa, London and Kitchener. Further, we see the scattering of private schools in the South East of Ontario, and then very few in the West of Ontario. While Southern Ontario is far more popular this relationship is notable as there is a growing Urban and Rural divide in Ontario. Furthermore, the distribution of private and public schools in Toronto is noteworthy because Toronto is not only the wealthiest city in Canada but also the most unequal with an increasing wealth gap between its rich and poor residents. To further address this distribution, I have graphed the distribution of school types across the Greater Toronto Area (gta).

```{r}
#| echo: false
#| warning: false
#| label: map_3

@@

#Plotting the map
ggmap(gta_map,
      extent = "normal",
      maprange = FALSE) +
  annotate("text", x = -79.7, y = 44., label = "Greater Toronto Area", size = 3, fontface = "bold", color = "black") +
  geom_point(data = ontario_gta_schools %>% filter(Facility_Type != "Private"), aes(x = Longitude, y = Latitude, color = Facility_Type), size = 1, alpha = 0.5) +
  geom_point(data = ontario_gta_schools %>% filter(Facility_Type == "Private"), aes(x = Longitude, y = Latitude, color = Facility_Type), size = 1, alpha = 0.5) +
  scale_color_manual(values = c("green", "red", "pink")) +
  labs(title = "Spatial Distribution of Types of Schools in the Greater Toronto Area", color = "Facility Type") +
  theme_minimal()

@map_3 highlights the spacial distribution of school types across the following cities in the GTA: Toronto, Mississauga, Brampton, Markham, Vaughn, Oakville, Burlington, Richmond Hill, Newmarket, Pickering, Ajax, Whitby and Oshawa. From this map, it is apparent that private schools are most heavily populated in the richest areas of Toronto, and as it moves area from Toronto, the number of public schools appears to dwindle. We see that public schools cluster most heavily around Toronto, Oakville and Richmond hill which are some of the wealthiest cities in Toronto. The cause of this distribution is likely a result of the differences in wealth within these cities.

@@

## Canadian Educational Disparities: A case study

https://www.theglobeandmail.com/news/national/time-to-lead/a-tale-of-two-schools-the-correlation-between-income-and-education/article15463950/ (cite 1)

https://www150.statcan.gc.ca/n1/pub/11f0019m/11f0019m2015367-eng.htm (cite 2)

https://www-tandfonline-com.myaccess.library.utoronto.ca/doi/full/10.1080/02680939.2017.1346203 (cite 3)

To better support the findings of this paper, I now turn my efforts toward a review of research on the educational disparities within Canada in relation to their geographical wealth. 
I have chosen to first focus my efforts on Ontario, as this was the main focus of my analysis. Toronto, Canada's largest city exhibits a growing divide between low and high-income individuals and thus exhibits a high wealth disparity. During an investigation into the high school education offered in two different neighbourhoods in Toronto The Globe and Mail argues that the difference between these schools, and the student's access to education is largely due to the socioeconomic backgrounds of the students who attend these schools (cite 1). One high school, Collegiate Institute is located in one of Toronto's wealthiest neighbourhoods, Forest Hill and the other,  Central Technical School, is located in Merish Village, a far less affluent area.  Students who attend education facilities in Forest Hill have higher academic achievement and see a greater amount of students pursuing post-secondary education and endeavours, Central Tech however does not showcase the same sentiments (cite 1). Students who reside in higher-income areas, such as forest hill have greater access to better educational resources in opposition to students in low-income families and areas who face far more barriers in regard to their access to quality education (cite 1). The Globe and Mail further discuss how these inequalities are perpetuated by the Canadian education system where schools in lower-income neighbourhoods receive less funding and resources than those of wealthier areas (cite 1).

To enhance my findings on geographical wealth and access to private and public schools I will now discuss a paper published by Statistic Canada titled Difference in Academic Performance Between Public and Private School Students by Geographic Region. Using the National Housing Survey to compare the academic performances of elementary to high school-aged students who attend public and private schools across Canada, their findings found that students attending private schools had higher academic performances than those in public schools (cite 2). Their findings were especially prominent in Atlantic provinces and Quebec (cite 2). The authors suggest that students who are able to attend private schools tend to come from high-income families and because of such, these types of private institutions are far more prevalent in high-income areas and neighbourhoods (cite 2).

A 2017 paper, titled The Geography of School Choice in a City with Growing Inequality: The Case of Vancouver sought to examine the socio-spatial impact of school choice on students' secondary school enrollment patterns (cite 3). Their investigation found that students who opt into over-subscribed schools (define this later) come from affluent areas and neighbourhoods with above-average levels of capital in Vancover (cite 3). Their findings found increasing levels of segregation among student mobility and the possibility that parents choose schools based on their socioeconomic characteristics within their affluent locations (cite 3). These finding indicate that schools in affluent neighbourhoods with dominant racial and ethnic groups are perceived as better than schools in less affluent neighborhoods (cite 3). → reword this

##Limitations and Bias 
These findings are in impart limited by the ODEF database from Statistics Canada, structural education policies, my manipulation and cleaning of variables, and biases imposed from my analysis.

>>>>

---
title: "From Coast to Coast: Uncovering the Provincial and Municipal impact of Wealth on the Spatial Distribution of Private and Public School Institutions in Canada"
author: "Sofia Sellitto"
thanks: "Code and data are available at: https://github.com/SofiaSellitto/cad_geo_schools"
date: "April 21, 2023"
date-format: long
format: 
  pdf: 
    documentclass: article 
    geometry: margin = 1in 
    abstract: "Inequitable access to education is a pervasive issue in the United States. However, Canada, a country with a similar education system, has received little attention in this area of research. This paper aims to  investigates the spatial and numerical distribution of educational facilities in Canada, particularly focused on private and public institutions. The study reveals that wealthier provinces like Ontario have a greater number and density of private schools, which has significant implications for students' access to quality education based on their geographical location and subsequent socio-economic status."
number-sections: true
output:
  bookdown::pdf_document2
toc: false
fig_caption: true
table_caption: true
nocite: '@'
bibliography: references.bib
---

```{r setup}
#| echo: false
#| warning: false
#| include: false

library(readr)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(janitor)
library(knitr)
library(ggmap)
library(gridtext)
library(gridExtra)
library(marginaleffects)

@@

data_cad <- read_csv(here::here("inputs/data/data.csv"))
```

```{r}
#| label: tbl-1
#| tbl-cap: Number of School Types Associated with Each Canadian Province
#| echo: false
#| warning: false

@@

@tbl-1 illustrates the numerical difference between each of the provinces of interest and their associated number of school types. It is apparent that Ontario maintains the largest number of schools total across all the Provinces of interest, with the greatest number of private institutions. Given that Ontario has both the largest population among all Canadian provinces and is the richest province with extreme wealth disparity, the observed numerical distribution appears to be accurate. Alberta maintains the second highest variation of public to private schools with Nova Scotia being the third.

\newpage
```{r}
#| label: fig-1
#| fig-cap: Number of School Types Associated with Each Canadian Province
#| echo: false
#| warning: false

@@

data_cad |>
  count(Prov_Terr, Facility_Type) |>
  ggplot(aes(x = Prov_Terr, y = n, fill = Facility_Type))+
  geom_col(position = "dodge")+
  labs(
    title = "Number of School Types for Each Province",
    x = "Province",
    y = "Number of School Types",
    fill = "School Type")+
  theme_minimal()+
  theme(
    plot.title = element_text(face="bold", size=14, hjust=0.5))
```

@fig-1 is a visual distribution of Table 1 to better aid in the comprehension of the data. Here the ratio of private to public schools is far more digestible. It also becomes apparent that Ontario appears to be the only Province with Catholic schools. This is due to the fact that Ontario is the only province that has a separate publicly funded Catholic school system. In other provinces and territories, these types of educational institutions would be classified as private schools and may thus add to the number of private schools for the the other, non-Ontario provinces.

```{r}
#| label: fig-2
#| fig-cap: Mapping the Spatial Distribution of School Types Across Canada
#| echo: false
#| warning: false

@@

 geom_point(data = data_cad %>% filter(Facility_Type != "Private"), 
            aes(x = Longitude, y = Latitude, color = Facility_Type), 
            size = 1, 
            alpha = 0.2
            ) +
geom_point(data = data_cad %>% filter(Facility_Type == "Private"), 
           aes(x = Longitude, y = Latitude, color = Facility_Type), 
           size = 1, alpha = 0.5
           ) +

  scale_color_manual(values = c("green", "orange", "purple", "red", "pink")) +

  labs(
    title = "Map of Canada",
    x = "Longitude",
    y = "Latitude",
    color = "School Type") +
  theme_minimal()+
  theme(
    plot.title = element_text(face="bold", size=14, hjust=0.5))

@fig-2 showcases the spatial distribution of school types across the map of Canada. In turning your attention toward the red dots, which denote public schools,  It is visually evident from the map that Private schools are concentrated in the highest volume in Ontario. Turning your attention to Alberta, it is evident that private institutions are far less concentrated in this province, and more heavily clustered in Northern Alberta, where Calgary would be. This highlights a potential disparity between rural and urban access to education. Next, we see that Nova Scotia has a lesser concentrated volume of private schools which would appear to be most concentrated in Halifax. Returning to the distribution of school types across Ontario, and given its status as the richest province with the highest income inequality, I have chosen to further examine the distribution and concentration of public and private institutions within the Ontario context.

```{r}
#| label: fig-3
#| fig-cap: Mapping the Spatial Distribution of School Types Across Ontario
#| echo: false
#| warning: false

@@

bbox <- c(left = -98, bottom = 41.80, right = -74.471, top = 52.92)

ontario_map <- get_stamenmap(bbox,
                             zoom = 7,
                             maptype = "terrain")
#Plotting the map 
ggmap(ontario_map,
      extent = "nomral",
      maprange = FALSE) +

geom_point(data = ontario_schools %>% filter(Facility_Type != "Private"), 
           aes(
             x = Longitude, 
             y = Latitude, 
             color = Facility_Type), 
             size = 1, alpha = 0.2
           ) +
  geom_point(data = ontario_schools %>% filter(Facility_Type == "Private"), 
             aes(
               x = Longitude, 
               y = Latitude, 
               color = Facility_Type), 
             size = 1, 
             alpha = 0.5
           ) +
scale_color_manual(values = c("green", "red", "pink")) +
  labs(
    title = "Map of Ontario",
    x = "Longitude",
    y = "Latitude",
    color = "School Type") +
  theme_minimal()+
  theme(
    plot.title = element_text(face="bold", size=14, hjust=0.5))
```

@fig-3 again is a spatial distribution of school types across the map of Ontario. Again using the same colours to denote the school types as in @fig-2, here it is apparent that the highest volume of private schools appears to cluster in Southern Ontario among the Greater Toronto Area (GTA), Ottawa, London and Kitchener. Further, we see a scattered clustering of private schools in the South East of Ontario, and then very few in the West of Ontario. While Southern Ontario is far more popular, this relationship is notable as there is a growing Urban and Rural divide in Ontario. Furthermore, the high concentration of private and public schools in Toronto is noteworthy as Toronto is not only the wealthiest city in Canada but also the most unequal with an increasing wealth gap between its rich and poor residents. To further address this distribution, I have graphed the distribution of school types across the Greater Toronto Area (gta).

```{r}
#| label: fig-4
#| fig-cap: Mapping the Spatial Distribution of School Types Across the Greater Toronto Area (GTA)
#| echo: false
#| warning: false

@@

#Plotting the map
ggmap(gta_map,
      extent = "normal",
      maprange = FALSE) +

  geom_point(data = ontario_gta_schools |> filter(Facility_Type != "Private"), 
             aes(
               x = Longitude, 
               y = Latitude,
               color = Facility_Type), 
             size = 1, 
             alpha = 0.2
             ) +
  geom_point(data = ontario_gta_schools |> filter(Facility_Type == "Private"), 
             aes(x = Longitude, 
                 y = Latitude, 
                 color = Facility_Type), 
             size = 1, 
             alpha = 0.2
             ) +

  scale_color_manual(values = c("green", "red", "pink")) +
  labs(
    title = "Map of GTA, Ontario",
    x = "Longitude",
    y = "Latitude",
    color = "School Type") +
  theme_minimal()+
  theme(
    plot.title = element_text(face="bold", size=14, hjust=0.5))

```
\newpage

@fig-4 highlights the spacial distribution of school types across the following cities in the GTA: Toronto, Mississauga, Brampton, Markham, Vaughn, Oakville, Burlington, Richmond Hill, Newmarket, Pickering, Ajax, Whitby and Oshawa. From this map, it is apparent that private schools are most heavily populated in the richest areas of Toronto, and as we move away from Toronto, the number of these institutions appears to dwindle. We see that public schools cluster most heavily around Toronto, Oakville and Richmond hill which are some of the wealthiest cities in Toronto. The cause of this distribution is likely a result of the differences in wealth within these cities.

@@

This paper has followed an analysis of the distribution and density of education facilities in the following Canadian Provinces: Ontario, Quebec, Alberta, British Columbia, Nova Scotia and New Brunswick, with the purpose of answering the following research objectives

(1) What is the distribution of educational facilities across Canada?

(2) Does access to public and private education differ spatially and numerically in accordance with Provincial and Municipal wealth?

It has paid particular attention to how provincial and municipal wealth impacts the density of private and public school institutions in Ontario and more particularly the Greater Toronto Area (GTA). Based on the data analysis conducted above, findings suggest that provinces that are wealthier, like Ontario, have a far greater number and density of private school institutions. When investigating the GTA, findings suggest that cities that are wealthier like Toronto, Oakville and Richmondhill, have a far greater density of private schools, in comparison to less wealthy cities in the GTA. These findings have implications on students' access to quality education and their possible failure to access education of a private school calibre as a result of their geographic location and subsequent SES due to their residence in such area. I will further discuss these findings in relation to the current literature in this field.

## Canadian Educational Disparities: A literature review

To better support the findings of this paper, I now turn my efforts toward a review of research on the educational disparities within Canada in relation to their geographical wealth. 
I have chosen to first focus my efforts on Ontario, as this was the main focus of my analysis. Toronto, the capital of Ontario and Canada's largest city exhibits a growing divide between low and high-income individuals and thus exhibits a high wealth disparity. During an investigation into the high school education offered in two different neighbourhoods in Toronto The Globe and Mail argues that the difference between these schools, and the student's access to education is largely due to the socioeconomic backgrounds of the students who attend these schools [@cite1]. One high school, Collegiate Institute is located in one of Toronto's wealthiest neighbourhoods, Forest Hill and the other,  Central Technical School, is located in Merish Village, a far less affluent area.  Students who attend education facilities in Forest Hill have higher academic achievement and see a greater amount of students pursuing post-secondary education and endeavours, Central Tech however does not showcase the same sentiments [@cite1]. Students who reside in higher-income areas, such as forest hill have greater access to better educational resources in opposition to students in low-income families and areas who face far more barriers in regard to their access to quality education [@cite1]. The Globe and Mail further discuss how these inequalities are perpetuated by the Canadian education system where schools in lower-income neighbourhoods receive less funding and resources than those of wealthier areas [@cite1].

To enhance my findings on geographical wealth and access to private and public schools I will now discuss a paper published by Statistic Canada titled Difference in Academic Performance Between Public and Private School Students by Geographic Region. Using the National Housing Survey to compare the academic performances of elementary to high school-aged students who attend public and private schools across Canada, their findings found that students attending private schools had higher academic performances than those in public schools [@cite2]. Their findings were especially prominent in Atlantic provinces and Quebec [@cite2]. The authors suggest that students who are able to attend private schools tend to come from high-income families and because of such, these types of private institutions are far more prevalent in high-income areas and neighbourhoods [@cite2].

A 2017 paper, titled The Geography of School Choice in a City with Growing Inequality: The Case of Vancouver sought to examine the socio-spatial impact of school choice on students' secondary school enrollment patterns [@cite3]. Their investigation found that students who opt into over-subscribed schools (define this later) come from affluent areas and neighbourhoods with above-average levels of capital in Vancover [@cite3]. Their findings found increasing levels of segregation among student mobility and the possibility that parents choose schools based on their socioeconomic characteristics within their affluent locations [@cite3]. These finding indicate that schools in affluent neighbourhoods with dominant racial and ethnic groups are perceived as better than schools in less affluent neighborhoods [@cite3]. → reword this

## Limitations and Bias

These findings are in impart limited by the ODEF database offered by Statistics Canada, my manipulation and cleaning of variables, and biases imposed from my analysis.
